Formalizing Stack Safety as a Security Property

### 2022-11-29 submission5/noexceptions.tex
Ready for sending

For confidentiality and integrity, our properties are {\em observational},
meaning that they allow a function activation to read from and write to other activations'
data, as long as such reads and writes do not affect the observable behavior of the
system. This permissiveness enables our properties to handle lazy enforcement mechanisms.

@@

We state our properties in terms of a generic machine subject to a few constraints,
but our examples and tests are focused on a RISC-V-like machine enhanced with PIPE,
a tag-based reference monitor. Importantly, such a machine has no explicit ``call''
instruction, only {\tt jal} instructions which may be used for that purpose or potentially
other purposes. Similarly, there is no ``return'', only {\tt jalr}, and so on.
We therefore define our properties under the assumption that we can distinguish the
steps that represent these abstract operations from those that do not---for instance,
by labeling instructions as calls, returns, etc.
Such a labeling scheme appears in
our testing environment (Section [TBD]) and is already necessary
to correctly apply tag-based enforcement in practice (cite something about this.)

@@

In section (TBD) we discuss the extension of this model into a simple concurrency setup.

@@

In addition to memory and registers, a machine has a program counter \(\PCname\)
and may have additional architecture-specific state; particularly relevant is state
that is used for security enforcement. We treat this as a set \(\mathcal{P}\) with
\(\PCname \in \mathcal{P}\). Collectively these are {\em state elements} \(\component\)
in the set \(\COMPONENTS ::= \PCname | \WORDS | \REGS | \mathcal{P}\).

@@

While some enforcement mechanisms may cause a machine to
failstop, we model this as the machine silently diverging by remaining in the
same state. Since failstops are not distinguished from normal divergence,
our properties will suffer the limitations typical of
{\it termination insensitive} systems \cite{}. TODO: more detailed explanation later.

@@

Suppose that {\tt f} is not a (valid) C function at all, but an attacker seeking
to leak {\tt secret}! It might do so in a number of ways, shown as snippets of
assembly code in Figure \ref{fig:f}.\footnote{We arrange that all the snippets
have the same length just to simplify some later explanations.}
%
In Figure \ref{subfig:direct}, {\tt f} takes an offset from the stack
pointer, accesses {\tt secret}, and directly outputs it. But more
subtly, even if somehow prevented from outputting {\tt secret} directly, {\tt f}
can instead return that value so that {\tt main} stores it to {\tt out},
as in Figure \ref{subfig:indirect}.
%
Beyond simply reading {\tt secret}, the attacker might overwrite {\tt sensitive}
with 42, guaranteeing that {\tt main} publishes its own secret unintentionally
(Figure \ref{subfig:integrity})!
Attacks of this kind do not violate {\tt main}'s confidentiality, but its
{\it integrity}.
And in Figure \ref{subfig:WBCF}, the attacker arranges to return to the
wrong instruction, thereby bypassing the check and publishing {\tt secret} regardless,
violating the program's {\it well-bracketed control flow} (WBCF.)
\apt{Add brief description of \ref{subfig:WBCF2}.}

@@

What if we want to express a finer-grained notion of safety, in which
such objects are protected unless the function that owns them intentionally
passes a pointer to them?

In order to express such a property, we need our machine to carry some notion
of {\it pointer provenance}---a distinction between a pointer that is intended to
point to a given object, and non-pointer integers as well as pointers to other objects.
Then \(\mathit{passed}\) should recursively identify all objects that
are reachable transitively from valid pointers in registers and accessible memory;
these keep their \(\object\) status, while those that are not reachable become \(\sealed\).

\section{Enforcement and Testing}

@@

I have a sketch of an argument that we would not gain anything meaningful from confidentiality
of tags, but it still needs some work. Still, I think it's correct not to vary the policy state.

>>>>

For confidentiality and integrity, our properties are {\em observational},
meaning that they are only concerned with the observable behavior of the system.
A function activation might read from and write to other activations' data
without violating an observational property, as long as such reads and writes do not
affect the observable behavior of the
system. This permissiveness enables our properties to handle lazy enforcement mechanisms.

@@

We state our properties in terms of a generic machine subject to a few constraints,
but our examples and tests are focused on a RISC-V-like machine enhanced with PIPE,
a tag-based reference monitor. Importantly, such a machine has no explicit ``call''
instruction, only {\tt jal} instructions which may be used for that purpose or potentially
other purposes. Similarly, there is no ``return'', only {\tt jalr}, and so on.
We therefore define our properties under the assumption that we can distinguish the
steps that represent these abstract operations from those that do not---for instance,
by labeling instructions as calls, returns, etc.
Such a labeling scheme appears in our testing environment (Section [TBD]).% and is already necessary
%to correctly apply tag-based enforcement in practice (cite something about this.)

@@

%In section (TBD) we discuss the extension of this model into a simple concurrency setup.

@@

In addition to memory and registers, a machine has a program counter \(\PCname\)
and may have additional architecture-specific state; particularly relevant is state
that is used for security enforcement. We treat this as a set \(\mathcal{P}\) with
\(\PCname \in \mathcal{P}\). Collectively these are {\em state elements} \(\component\)
in the set \(\COMPONENTS ::= \PCname | \WORDS | \REGS | \mathcal{P}\).
One instance of \(\mathcal{P}\) is discussed in Section \ref{sec:enforcement}.

@@

While some enforcement mechanisms may cause a machine to
failstop, we model this as the machine silently diverging by remaining in the
same state. We consider silently diverging traces to be indistinguishable
(from the point of view of an attacker) from traces that might eventually
produce output, an assumption that gives our properties the flavor of
{\it termination insensitivity}.

@@

Suppose that {\tt f} is not a (valid) C function at all, but an attacker seeking
to leak {\tt secret}! It might do so in a number of ways, shown as snippets of
assembly code in Figure \ref{fig:f}.\footnote{We arrange that all the snippets
have the same length just to simplify some later explanations.}
%
In Figure \ref{subfig:direct}, {\tt f} takes an offset from the stack
pointer, accesses {\tt secret}, and directly outputs it. But more
subtly, even if somehow prevented from outputting {\tt secret} directly, {\tt f}
can instead return that value so that {\tt main} stores it to {\tt out},
as in Figure \ref{subfig:indirect}.
%
Beyond simply reading {\tt secret}, the attacker might overwrite {\tt sensitive}
with 42, guaranteeing that {\tt main} publishes its own secret unintentionally
(Figure \ref{subfig:integrity})!
Attacks of this kind do not violate {\tt main}'s confidentiality, but its
{\it integrity}.
In Figure \ref{subfig:WBCF}, the attacker arranges to return to the
wrong instruction, thereby bypassing the check and publishing {\tt secret} regardless,
violating the program's {\it well-bracketed control flow} (WBCF.) Figure \ref{subfig:WBCF2}
also shows an attack that violates WBCF, this time by returning to the correct
program counter but with the stack pointer offset.

@@

What if we want to express a finer-grained notion of safety, in which
stack objects are protected unless the function that owns them intentionally
passes a pointer to them? This can be thought of as a {\it capability}-based
notion of security. Capabilities are unforgeable tokens that grant access to
a region of memory, typically corresponding to valid pointers to that region.
So, in order to express such a property, we need our machine to carry some notion
of {\it pointer provenance}---a distinction between a pointer that is intended to
point to a given object, and non-pointer integers as well as pointers to other objects.

One such provenance model is Memarian et al.'s PVI \cite{}, in which pointers are
annotated with the identity of the object they first pointed to. This annotation is
propagated when the pointer is copied and when operations are performed on it---even
integer-only operations.

We can model this as a pair of additional security-relevant operations: one which
declares a register to contain a valid pointer, and one which transmits the provenance
of a pointer from one element to another.

\begin{align*}
  \psi \in \Psi ::= & \dots \\
  & \mathbf{promote} ~ \reg ~ sa & sa \in \REGS \times \mathbb{Z} \times \mathbb{N} \\
  & \mathbf{propagate} ~ \component_{src} ~ \component_{dst} \\
\end{align*}

In addition to the normal call stack, our security context will carry a map \(\rho\) from
elements to memory regions, represented as a base and a bound \(\context = (V, \sigma, \rho)\).
Existing operations are extended to propagate \(\rho\), and the new operations work as follows:

\judgmenttwobrlong[]
                  {\(\psi = \mathbf{promote} ~ \reg_{dst} ~ (\reg_{base},\mathit{off},sz)\)}
                  {\(b = \mach[\reg_{base}]+\mathit{off}\)}
                  {\(\rho' = \rho[\reg \mapsto (b,b+sz)]\)}
                  {\(Op ~ \mach ~ \psi ~ (V,\sigma,\rho) = (V,\sigma,\rho')\)}

\judgmenttwo[]
            {\(\psi = \mathbf{propagate} ~ \component_{src} ~ \component_{dst}\)}
            {\(\rho' = \rho[\component_{dst} \mapsto \rho[\component_{src}]]\)}
            {\(Op ~ \mach ~ \psi ~ (V,\sigma,\rho) = (V,\sigma,\rho')\)}

We now have a notion of provenance, and must integrate it into the definition of
stack safety. We essentially generalize the above notion of passing: we will consider
a caller to have intentionally passed an object if that object is reachable by
a capability that has been passed to the callee. This includes capabilities passed
indirectly, by being stored in an object that is in turn passed. Formally, we call
this set \(\mathit{capped}\), and define it recursively:

\[\mathit{capped} ~ \components ~ \rho \triangleq
\bigcup_{\component \in \components} \{\component' |
\rho[\component] = (\mathit{base},\mathit{bound})
\land \mathit{base} \leq \component' < \mathit{bound}\}\]

We then tweak the call operation to seal only objects that are in \(\mathit{capped}\), or
the previously defined \(\mathit{passed}\).

\judgmentbrbrbr[]
               {\(\psi = \mathbf{call} ~ \addr_{target} ~ \overline{\reg_{args}} ~ \overline{sa}\)}
               {\(V' = V \llbracket \reg \mapsto \unsealed | \reg \in \mathit{CLR} \rrbracket
                 \llbracket \reg \mapsto \public | \reg \in \overline{\reg_{args}} \rrbracket\)}
               {\(\components = \{\addr | V' ~ \addr = \object \land \addr \not \in (\mathit{passed} ~ \overline{sa} ~ \mach) \cup (\mathit{capped} ~ \overline{\reg_{args}} ~ \rho) \}\)}
               {\(V'' = V'\llbracket \components \mapsto \sealed \rrbracket\)}
               {\(Op ~ \mach ~ \psi ~ (V,\sigma,\rho) =
                 (V'',V::\sigma,\rho)\)}

Note that we have a degree of monotonicity here. Once an object is sealed (because its
capability has not been passed to a callee), subsequent nested calls can never unseal it.
On the other hand, an object that is passed via a pointer may be passed on indefinitely.

\section{Enforcement}
\label{sec:enforcement}

@@

\section*{Remainder}

These document sections lay out a way to present our stack safety properties to build intuition and
then introduce the formalism. Next the full paper will need the testing section and an expanded
enforcement section, but these can largely be borrowed from previous submissions.
